The evolution of gene expression in the term placenta of viviparous mammals

### 2016-04-06 manuscript/manuscript.Rnw
clean up manuscript and supplemental material

%\documentclass[english,12pt]{article}
\documentclass[doublespacing]{bmcart_dla}
\usepackage{fontspec}
\setmainfont{FreeSans}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSans}
\setmathsf{FreeSans}
\setmathtt{FreeSans}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage[letterpaper]{geometry}
\usepackage{url}
% \usepackage{fancyhdr}
\usepackage{graphicx}
% \usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
%\usepackage[plain]{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage{sfmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
% \usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
\usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
% \newenvironment{narrow}[2]{%
%   \begin{list}{}{%
%       \setlength{\topsep}{0pt}%
%       \setlength{\leftmargin}{#1}%
%       \setlength{\rightmargin}{#2}%
%       \setlength{\listparindent}{\parindent}%
%       \setlength{\itemindent}{\parindent}%
%       \setlength{\parsep}{\parskip}}%
%   \item[]}{\end{list}}
% \def\newblock{\hskip}
% \newenvironment{paperquote}{%
%   \begin{quote}%
%      \it
%   }%
%   {\end{quote}}
% \renewcommand{\textfraction}{0.15}
% \renewcommand{\topfraction}{0.85}
% \renewcommand{\bottomfraction}{0.65}
% \renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
% \newenvironment{enumerate*}%
%   {\begin{enumerate}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{enumerate}}
% \newenvironment{itemize*}%
%   {\begin{itemize}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{itemize}}
% \setcounter{tocdepth}{3}
% \oddsidemargin 0.0in 
% \textwidth 6in
% \raggedbottom
% \clubpenalty = 10000
% \widowpenalty = 10000
% set a large emergency stretch to avoid issues with overfull boxes
% \setlength{\emergencystretch}{20em}
% make the afill smaller
% \renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
% \pagestyle{fancy}

@@

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of invasiveness of fetal tissues into maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(\cref{tab:gestational_periods}).

@@

% \DLA{Add details for new sequences and fix numbers}

@@

For more complete evolutionary comparisons additional sequences of
placenta in \mouse
(n=\Sexpr{star.logs[grepl("musculus",species),length(unique(file))]})
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
(n=\Sexpr{star.logs[grepl("sapiens",species),length(unique(file))]})
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
\ac{SRA} (Supplemental \cref{tab:srr_numbers}) and analyzed using the
same methodology as mRNAs sequenced specifically for this study.

@@

All of the code used to produce every analysis presented in this paper
is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean FPKM for gene in each
species with ortholog annotations are located in the supplemental data
files \Sexpr{english.join(gsub("_","\\\\_",paste0(gsub(" ","_",
  star.logs[,unique(species)]),"_mean_fpkm.txt")))}.

@@

@@

        display.table <-
            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
                                                  "}")]
        setnames(display.table,"Fold Change","$\\mathrm{log}_2$ FC")
        print(xtable(display.table,
                     display=c("s","s","s","e","g","f"),
                     align=c("c","c","p{2in}","r","r","r")
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

@@

We examined the expression of three primates, \human, \bonobo, and
\smonkey. The genes whose expressions are different on the primate
lineage in comparison to all other analyzed species are listed in
\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
higher expression in the primate lineage
(Supplemental \cref{fig:fstl1_evolution_tree},
$\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
in BMP signaling, and has previously been implicated in inflammation
and invasiveness
\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
non-primates, \horse shows the highest expression
%
\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
%
which is still lower than the lowest primate, \smonkey
%
(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}). 
%
No other genes were significantly differentially expressed on this
lineage by t-test.

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \mouse are shown in
\cref{tab:top50_mouse_placenta} with a list of all genes with
$\mathrm{FPKM} \ge 50$ in Supplemental
\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
target \textit{in silico} is the orphan G-coupled receptor
\textit{Mrgprg}
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
of which are maternally imprinted (the later of which is a homolog of
human \textit{H19\_2}),
\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
cathepsins and a rodentia specific expansion of prolactin related
genes. There are also two miRNA which overlap with the of the 7S
signal recognition particle
(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
and
\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
and numerous pseudogenes
(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
appear to be highly expressed, however, the short read length (34--44)
in many samples is likely preventing unambiguous assignment of all
reads to the more lengthy complete transcript, so these shorter
transcripts should be treated with suspicion, and we will ignore them
in further discussion. The prolactin-related genes
(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
very highly expressed in \mouse in, with %
\Sexpr{combined.fpkm[species=="mus musculus" &
  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
prolactin related genes having $\mathrm{FPKM} \ge 1$
(\cref{fig:prolactin_evolution_tree}, Supplemental
\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
also expanded in rodents, with six showing high expression
(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
\cref{tab:top50_mouse_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
Finally, \textit{Peg10}, an imprinted gene required for proper
formation of the placenta in mice is also highly expressed
(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.

@@

The top 50 expressed genes in the term placenta of \armadillo are
shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
with $\mathrm{FPKM}\ge 50$ in Supplemental
\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
highest expression is
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
which is most similar to the \mouse \textit{Pga5} is expanded in
\opossum into three members which are expressed in placenta (the other
two being
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
and
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
though the expansion is less than that seen \cow and \sheep
(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). There is also high expression of
hemoglobin subunit beta,
\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
an un-annotated gene with high similarity to \human and \cow
\textit{OVOS2}
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
(Supplemental \cref{fig:ovos2_evolution_tree}), the growth factors
\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
and
\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
and the thymosin
\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
As has been seen previously, multiple components of the annexin
complex are also highly expressed, including
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
(\textit{ANXA8L1}),
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
(\textit{S100B}),
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
(\textit{S100A10}), and
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
(\textit{ANXA2}).

>>>>

%\documentclass[english,12pt]{article}
\documentclass[doublespacing]{bmcart_dla}
\usepackage{fontspec}
\setmainfont{FreeSans}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSans}
\setmathsf{FreeSans}
\setmathtt{FreeSans}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage[letterpaper]{geometry}
\usepackage{url}
% \usepackage{fancyhdr}
\usepackage{graphicx}
% \usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
%\usepackage[plain]{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{amsmath}
% \usepackage{sfmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
% \usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
\usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
% \newenvironment{narrow}[2]{%
%   \begin{list}{}{%
%       \setlength{\topsep}{0pt}%
%       \setlength{\leftmargin}{#1}%
%       \setlength{\rightmargin}{#2}%
%       \setlength{\listparindent}{\parindent}%
%       \setlength{\itemindent}{\parindent}%
%       \setlength{\parsep}{\parskip}}%
%   \item[]}{\end{list}}
% \def\newblock{\hskip}
% \newenvironment{paperquote}{%
%   \begin{quote}%
%      \it
%   }%
%   {\end{quote}}
% \renewcommand{\textfraction}{0.15}
% \renewcommand{\topfraction}{0.85}
% \renewcommand{\bottomfraction}{0.65}
% \renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
% \newenvironment{enumerate*}%
%   {\begin{enumerate}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{enumerate}}
% \newenvironment{itemize*}%
%   {\begin{itemize}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{itemize}}
% \setcounter{tocdepth}{3}
% \oddsidemargin 0.0in 
% \textwidth 6in
% \raggedbottom
% \clubpenalty = 10000
% \widowpenalty = 10000
% set a large emergency stretch to avoid issues with overfull boxes
% \setlength{\emergencystretch}{20em}
% make the afill smaller
% \renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
% \pagestyle{fancy}

@@

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe,Moffett.ea2006:Immunologyofplacentationin}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of invasiveness of fetal tissues into maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(\cref{tab:gestational_periods}).

@@

@@

For more complete evolutionary comparisons additional sequences of
placenta in \mouse
(n=\Sexpr{star.logs[grepl("musculus",species),length(unique(file))]})
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
(n=\Sexpr{star.logs[grepl("sapiens",species),length(unique(file))]})
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
\ac{SRA} (Supplemental \cref{tab:srr_numbers}) and analyzed using the
same methodology as RNAs sequenced specifically for this study.

@@

All of the code used to produce every analysis presented in this paper
is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean FPKM for gene in each
species with ortholog annotations are located in the supplemental data
file all\_species\_mean\_fpkm.txt.xz and the per-sample FPKM values are in
all\_species\_per\_sample\_fpkm.txt.xz.

@@

    if (is.na(fpkm)) {
        stop("FPKM is NA!!!!")

@@

        display.table <-
            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
                                                  "}")]
        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
        print(xtable(display.table,
                     display=c("s","s","s","e","g","f"),
                     align=c("c","c","p{2in}","r","r","r")
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

@@

We examined the expression of three primates, \human, \bonobo, and
\smonkey. The genes whose expressions are different on the primate
lineage in comparison to all other analyzed species are listed in
\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
higher expression in the primate lineage
(Supplemental \cref{fig:fstl1_evolution_tree},
$\text{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
in BMP signaling, and has previously been implicated in inflammation
and invasiveness
\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
non-primates, \horse shows the highest expression
%
\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
%
which is still lower than the lowest primate, \smonkey
%
(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}). 
%
No other genes were significantly differentially expressed on this
lineage by t-test.

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \mouse are shown in
\cref{tab:top50_mouse_placenta} with a list of all genes with
$\text{FPKM} \ge 50$ in Supplemental
\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
target \textit{in silico} is the orphan G-coupled receptor
\textit{Mrgprg}
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
of which are maternally imprinted (the later of which is a homolog of
human \textit{H19\_2}),
\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
cathepsins and a rodentia specific expansion of prolactin related
genes. There are also two miRNA which overlap with the of the 7S
signal recognition particle
(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
and
\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
and numerous pseudogenes
(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
appear to be highly expressed, however, the short read length (34--44)
in many samples is likely preventing unambiguous assignment of all
reads to the more lengthy complete transcript, so these shorter
transcripts should be treated with suspicion, and we will ignore them
in further discussion. The prolactin-related genes
(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
very highly expressed in \mouse in, with %
\Sexpr{combined.fpkm[species=="mus musculus" &
  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
prolactin related genes having $\text{FPKM} \ge 1$
(\cref{fig:prolactin_evolution_tree}, Supplemental
\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
also expanded in rodents, with six showing high expression
(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
\cref{tab:top50_mouse_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
Finally, \textit{Peg10}, an imprinted gene required for proper
formation of the placenta in mice is also highly expressed
(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.

@@

The top 50 expressed genes in the term placenta of \armadillo are
shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
with $\text{FPKM}\ge 50$ in Supplemental
\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
highest expression is
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
which is most similar to the \mouse \textit{Pga5} is expanded in
\opossum into three members which are expressed in placenta (the other
two being
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
and
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
though the expansion is less than that seen \cow and \sheep
(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). There is also high expression of
hemoglobin subunit beta,
\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
an un-annotated gene with high similarity to \human and \cow
\textit{OVOS2}
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
(Supplemental \cref{fig:ovos2_evolution_tree}), the growth factors
\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
and
\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
and the thymosin
\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
As has been seen previously, multiple components of the annexin
complex are also highly expressed, including
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
(\textit{ANXA8L1}),
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
(\textit{S100B}),
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
(\textit{S100A10}), and
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
(\textit{ANXA2}).
